The helicoverpa package

### 2018-08-14 notebook/helicoverpa_notes.Rmd
Fix link
Update notebook

     - **AUG** [13](#u-13-AUG-2018)
     - **JUL** [18](#u-18-JUL-2018), [23](#u-23-JUL-2018), [30](#u-30-JUL-2018)

@@

> <a name="u-18-AUG-2018">Update: 18 AUG 2018</a>

@@

>>>>

     - **AUG** [13](#u-13-AUG-2018), [14](#u-14-AUG-2018)
     - **JUL** [18](#u-18-JUL-2018), [23](#u-23-JUL-2018), [30](#u-30-JUL-2018)

@@

> <a name="u-14-AUG-2018">Update: 14 AUG 2018</a>

```{r, echo = FALSE}
toy_simulate_resistance <- function(generations = 20,       # Generations to sim
                                    xdim = 2,               # Land dimension 1
                                    ydim = 2,               # Land dimension 2
                                    pathogens = 1,          # Pathogen strains
                                    crops = 1,              # Crop species
                                    path_alleles = 3,       # Pathogen alleles
                                    crop_alleles = 3,       # Crop alleles
                                    pest_init = 2000,       # Initial pests 
                                    crop_rotate = "static", # Crops rotated
                                    path_rotate = "static", # Pathogens rotated
                                    pest_move_pr = 0.1,     # Pest movement
                                    pest_move_dist = 1,     # Pest move distance
                                    fecundity = 8,          # Offspring per fem
                                    cell_K = 2000           # K per cell
                                    ){

    if(pest_move_dist > xdim & pest_move_dist > ydim){
        pest_move_dist <- max(c(xdim, ydim)); # Avoids error
    }
    # Start initialising the landscape and pests
    LAND <- toy_initialise_land(xdim  = xdim, ydim = ydim, 
                                pathogens = pathogens, crops = crops);
    PEST <- toy_initialise_pest(LAND, N = pest_init, p_al = path_alleles, 
                                c_al = crop_alleles);
    # Start the generations
    PEST_DATA   <- NULL;
    gen         <- 1;
    while(gen < generations){
        LAND <- toy_set_crops(LAND, crops, crop_rotate);                        
        LAND <- toy_set_paths(LAND, pathogens, path_rotate);                    
        PEST <- toy_move_pest(PEST, LAND, pest_move_pr, pest_move_dist);        
        PEST <- toy_feed_pest(PEST, LAND);                                      
        if(toy_check_extinction(PEST, gen) == TRUE){ # Hate these if breaks here
            break;
        }
        PEST <- toy_kill_pest(PEST, LAND);
        if(toy_check_extinction(PEST, gen) == TRUE){
            break;
        }
        PEST <- toy_reproduce_pest(PEST, LAND, path_alleles, crop_alleles, 
                                   fecundity, cell_K);
        if(toy_check_extinction(PEST, gen) == TRUE){
            break;
        }
        PEST_DATA[[gen]] <- PEST;
        gen <- gen + 1;
    }
    return(PEST_DATA);
}

summarise_pest_data <- function(PEST_DATA){
    # Density estimates
    den_list  <- unlist(lapply(PEST_DATA, dim));
    den_vect  <- den_list[c(TRUE, FALSE)];
    gens      <- length(PEST_DATA);
    p_alleles <- max(c(PEST_DATA[[1]][,5], PEST_DATA[[1]][,6]));
    c_alleles <- max(c(PEST_DATA[[1]][,7], PEST_DATA[[1]][,8])); 
    p_tabl    <- matrix(data = 0, nrow = gens, ncol = p_alleles);
    c_tabl    <- matrix(data = 0, nrow = gens, ncol = c_alleles);
    # Allele frequencies
    for(gen in 1:length(PEST_DATA)){
        total_p_alleles <- length(PEST_DATA[[gen]][,5:6]);
        total_c_alleles <- length(PEST_DATA[[gen]][,7:8]);
        for(allele in 1:p_alleles){
            allele_count         <- sum(PEST_DATA[[gen]][,5:6] == allele);
            p_tabl[gen, allele]  <- allele_count / total_p_alleles
        }
        for(allele in 1:c_alleles){
            allele_count         <- sum(PEST_DATA[[gen]][,7:8] == allele);
            c_tabl[gen, allele]  <- allele_count / total_c_alleles
        }
    }
    return(list(densities = den_vect, pathogen_fr = p_tabl, crop_fr = c_tabl));
}

**More playing around with the toy model**

Following up on [yesterday](#skip_AUG_code), I have tweaked two functions to allow us to manipulate the number of alleles underlying pathogen resistance and crop feeding. In effect, this allows us to see how selection can sweep through to nearly fix an allele for resistance to a particular pathogen or for feeding on a particular crop (since only one allele is needed, other neutral alleles can hide in the heterozygotes). But, when crops are rotated, or pathogens are rotated, we have higher genetic diversity -- note that we are making an implicit assumption of a strong negative genetic correlation; pests can only have two alleles for resistance, e.g., and if there are three pathogens, then all pests must necessarily be susceptible to 1-2 pathogens.

```{r}
# ==============================================================================
# This is just a toy version of the model, for the presentation
# ==============================================================================
# Note that I'm adding 'toy' at the start of each function; this is just in case
# we actually want to keep some of these functions in the eventual R package, 
# but also don't want to worry about not being able to name functions like
# "initialise_land" the same here as elsewhere in the eventual bigger model.
# ==============================================================================
toy_simulate_resistance <- function(generations = 20,       # Generations to sim
                                    xdim = 2,               # Land dimension 1
                                    ydim = 2,               # Land dimension 2
                                    pathogens = 1,          # Pathogen strains
                                    crops = 1,              # Crop species
                                    path_alleles = 3,       # Pathogen alleles
                                    crop_alleles = 3,       # Crop alleles
                                    pest_init = 2000,       # Initial pests 
                                    crop_rotate = "static", # Crops rotated
                                    path_rotate = "static", # Pathogens rotated
                                    pest_move_pr = 0.1,     # Pest movement
                                    pest_move_dist = 1,     # Pest move distance
                                    fecundity = 8,          # Offspring per fem
                                    cell_K = 2000           # K per cell
                                    ){

    if(pest_move_dist > xdim & pest_move_dist > ydim){
        pest_move_dist <- max(c(xdim, ydim)); # Avoids error
    }
    # Start initialising the landscape and pests
    LAND <- toy_initialise_land(xdim  = xdim, ydim = ydim, 
                                pathogens = pathogens, crops = crops);
    PEST <- toy_initialise_pest(LAND, N = pest_init, p_al = path_alleles, 
                                c_al = crop_alleles);
    # Start the generations
    PEST_DATA   <- NULL;
    gen         <- 1;
    while(gen < generations){
        LAND <- toy_set_crops(LAND, crops, crop_rotate);                        
        LAND <- toy_set_paths(LAND, pathogens, path_rotate);                    
        PEST <- toy_move_pest(PEST, LAND, pest_move_pr, pest_move_dist);        
        PEST <- toy_feed_pest(PEST, LAND);                                      
        if(toy_check_extinction(PEST, gen) == TRUE){ # Hate these if breaks here
            break;
        }
        PEST <- toy_kill_pest(PEST, LAND);
        if(toy_check_extinction(PEST, gen) == TRUE){
            break;
        }
        PEST <- toy_reproduce_pest(PEST, LAND, path_alleles, crop_alleles, 
                                   fecundity, cell_K);
        if(toy_check_extinction(PEST, gen) == TRUE){
            break;
        }
        PEST_DATA[[gen]] <- PEST;
        gen <- gen + 1;
    }
    return(PEST_DATA);
}

summarise_pest_data <- function(PEST_DATA){
    # Density estimates
    den_list  <- unlist(lapply(PEST_DATA, dim));
    den_vect  <- den_list[c(TRUE, FALSE)];
    gens      <- length(PEST_DATA);
    p_alleles <- max(c(PEST_DATA[[1]][,5], PEST_DATA[[1]][,6]));
    c_alleles <- max(c(PEST_DATA[[1]][,7], PEST_DATA[[1]][,8])); 
    p_tabl    <- matrix(data = 0, nrow = gens, ncol = p_alleles);
    c_tabl    <- matrix(data = 0, nrow = gens, ncol = c_alleles);
    # Allele frequencies
    for(gen in 1:length(PEST_DATA)){
        total_p_alleles <- length(PEST_DATA[[gen]][,5:6]);
        total_c_alleles <- length(PEST_DATA[[gen]][,7:8]);
        for(allele in 1:p_alleles){
            allele_count         <- sum(PEST_DATA[[gen]][,5:6] == allele);
            p_tabl[gen, allele]  <- allele_count / total_p_alleles
        }
        for(allele in 1:c_alleles){
            allele_count         <- sum(PEST_DATA[[gen]][,7:8] == allele);
            c_tabl[gen, allele]  <- allele_count / total_c_alleles
        }
    }
    return(list(densities = den_vect, pathogen_fr = p_tabl, crop_fr = c_tabl));
}
```

I'm going to run the same analysis as [yesterday](#skip_AUG_code), but now show how allele frequencies change over generations. I've not broken this down by landscape cell yet, but this will be easy to do (just let me know the format the data would be best presented in). In each case, now there will be 3 alleles underlying pathogen resistance, and 3 alleles underlying crop feeding.

```{r}
sim1 <- toy_simulate_resistance(pathogens = 1, crops = 1, cell_K = 2000, 
                                pest_init = 2000, fecundity = 8, 
                                generations = 20, pest_move_pr = 0.1, 
                                crop_rotate = "static", path_rotate = "static");
dat1 <- summarise_pest_data(sim1);
plot(x = 1:length(dat1$densities), y = dat1$densities, type = "b", pch = 20, 
     lwd = 2, cex.lab = 1.5, cex.axis = 1.5, ylim = c(0, 8000), xlim = c(1, 19),
     xlab = "Generation", ylab = "Pest density");
plot(x = 1:length(dat1$pathogen_fr[,1]), y = dat1$pathogen_fr[,1], type = "b", 
     pch = 20, lwd = 2, cex.lab = 1.5, cex.axis = 1.5, ylim = c(0, 1), 
     xlim = c(1, 19), xlab = "Generation", ylab = "Allele frequency");
points(x = 1:length(dat1$pathogen_fr[,2]), y = dat1$pathogen_fr[,2], type = "b", 
     pch = 20, lwd = 2, col = "red");
points(x = 1:length(dat1$pathogen_fr[,3]), y = dat1$pathogen_fr[,3], type = "b", 
     pch = 20, lwd = 2, col = "blue");
```

```{r}
sim2 <- toy_simulate_resistance(pathogens = 3, crops = 1, cell_K = 2000, 
                                pest_init = 2000, fecundity = 8, 
                                generations = 20, pest_move_pr = 0.1, 
                                crop_rotate = "static", path_rotate = "random");
dat2 <- summarise_pest_data(sim2);
plot(x = 1:length(dat2$densities), y = dat2$densities, type = "b", pch = 20, 
     lwd = 2, cex.lab = 1.5, cex.axis = 1.5, ylim = c(0, 8000), xlim = c(1, 19),
     xlab = "Generation", ylab = "Pest density");
plot(x = 1:length(dat2$pathogen_fr[,1]), y = dat2$pathogen_fr[,1], type = "b", 
     pch = 20, lwd = 2, cex.lab = 1.5, cex.axis = 1.5, ylim = c(0, 1), 
     xlim = c(1, 19), xlab = "Generation", ylab = "Allele frequency");
points(x = 1:length(dat2$pathogen_fr[,2]), y = dat2$pathogen_fr[,2], type = "b", 
     pch = 20, lwd = 2, col = "red");
points(x = 1:length(dat2$pathogen_fr[,3]), y = dat2$pathogen_fr[,3], type = "b", 
     pch = 20, lwd = 2, col = "blue");
```

Next, we can see what happens when we instead have 3 pathogens *and* 3 crops under otherwise identical conditions. In each generation, one of three pathogens and one of three crops are randomly applied to each landscape cell.

```{r}
sim3 <- toy_simulate_resistance(pathogens = 3, crops = 3, cell_K = 2000, 
                                pest_init = 2000, fecundity = 8, 
                                generations = 20, pest_move_pr = 0.1, 
                                crop_rotate = "random", path_rotate = "random");
dat3 <- summarise_pest_data(sim3);
plot(x = 1:length(dat3$densities), y = dat3$densities, type = "b", pch = 20, 
     lwd = 2, cex.lab = 1.5, cex.axis = 1.5, ylim = c(0, 8000), xlim = c(1, 19),
     xlab = "Generation", ylab = "Pest density");
plot(x = 1:length(dat3$pathogen_fr[,1]), y = dat3$pathogen_fr[,1], type = "b", 
     pch = 20, lwd = 2, cex.lab = 1.5, cex.axis = 1.5, ylim = c(0, 1), 
     xlim = c(1, 19), xlab = "Generation", ylab = "Allele frequency");
points(x = 1:length(dat3$pathogen_fr[,2]), y = dat3$pathogen_fr[,2], type = "b", 
     pch = 20, lwd = 2, col = "red");
points(x = 1:length(dat3$pathogen_fr[,3]), y = dat3$pathogen_fr[,3], type = "b", 
     pch = 20, lwd = 2, col = "blue");
```

As is evident, pathogen and crop rotation decreases pest density and increases the diversity of alleles underlying pathogen resistance and crop feeding.

> <a name="u-13-AUG-2018">Update: 13 AUG 2018</a>

@@

Next, we can see what happens when we instead have 3 pathogens *and* 3 crops under otherwise identical conditions. In each generation, one of three pathogens and one of three crops are randomly applied to each landscape cell.
